|

### 2021-10-08 05-Chp-CPRD-Analysis.Rmd
End of day:

```{r, echo = FALSE, warning=FALSE, message=FALSE}
source("R/doc_options.R")
source("R/helper.R")
knitr::read_chunk("R/Chp3-CPRD-Analysis.R")
doc_type <- knitr::opts_knit$get('rmarkdown.pandoc.to') # Info on knitting format
```

## Additional ideas

One of the problems I'll point out with the trials in the systematic review chapter (Chapter \@ref(sys-rev-heading)) is that they only included people with a high cardiovascular risk - we are kind of doing the same by using elevated test cholesterol results as the index event.

Might be able to get around this by pointing out that you need some index event to be able to do this analysis (?)

__Things we tried__

- Conditioning entry to the cohort on QRISK score (as defined by codes) rather than by . Need to be able to demonstrate here that not many meeting our original entry criteria actually go on to have a statin quickly. Need total number of those starting and average time to start.
- Positive controls. Need to be able to explain why these controls have such wildly increased results. Cross-reference these results with the Hernan paper.
- Competing risk analysis, with death as a competing risk. Need numbers of deaths in each group, and preliminary results
- Allowing for time-varying confounders for binary covariates.

__Things we could try__

- Work through data using positive control to see if the way it is being added is what is causing the issues.
- Examine Cramer et al. thesis and see what they did
- Marginal structural models approach (a la J. Sterne)

<!----------------------------------------------------------------------------->

## Lay summary
Large scale database such as this are often called population level, and studies that make use of them have a number of advantages.

This analysis uses one of these databases, the Clinical Practice Research Datalink (CPRD), which contains the electronic medical records of XXXXXX people from YYYYYY general practices across England.

Using this data, I have analysed the ...

<!----------------------------------------------------------------------------->

@@

Lack of evidence for the association of statins on vascular dementia risk

Large scale databases advantages over other means

Cross reference here aswell re: the lack of information on Vascular dementia and the need for analyses at different sources of bias.

<!----------------------------------------------------------------------------->

@@

@@

The Clinical Practice Research Datalink (CPRD) is a large, population-based, electronic health record database.[@herrett2015] It contains the primary care records for

Originally termed the GPRD (General Practice Research Database),

The

Data completeness and collection of certain disease are heavily influenced by exercises that reward doctors for capturing certain diagnoses, such as the Quality and Outcomes Framework. However, these drives can oftern have unintended consequences [@liu2019]

The CPRD uses a coding system called Read code <!--- check capitalisation --->

### Study design and protocol

We performed a prospective cohort study using data from the CPRD. Our initial sample included all participants registered at a participating practice between 1 January 1995 and 29 February 2016 who had a flag for "research quality" data. All events of interest were identified using predetermined code lists, which are available for inspection (see [Data/code availability](#data-code-avail)).

An *a priori* protocol for this study was published, [@walker2016a] and amendments to this are recorded in Supplementary Materials 1. This study was reported in line with the STROBE Cohort guidelines (Supplementary Table 3).[@vonelm2008]

<!-- Note: If protocol was written by Venexia, how do I reference this? -->

@@

### Study cohort

Participants were included in our study cohort if their record contained any of the following index events: a Read code for a diagnosis of hypercholesterolemia or related condition; a Read code for prescription of a lipid-regulating agent (such as statins); a total cholesterol test result of >4 mmol/L; or an LDL-c test result of >2 mmol/L.

These index events allowed us to define a population of participants who were either at risk of hypercholesterolemia, as indicated by the elevated total or LDL cholesterol test results, or had already been diagnosed with it, as indicated by a diagnostic code/related prescription. This approach was employed in an attempt to reduce confounding by indication that we would expect to observe in the full cohort, because individuals not prescribed lipid-regulating agents likely be less healthy across a range of variables than those prescribed lipid-regulating agents, leading to a biased association been lipid-regulating agent use and dementia. Conditioning entry into the study into the study on being either “at-risk” or already diagnosed with hypercholesterolemia attempts to mitigate this bias.

The index date for a participant was defined as the date where the first relevant code or test result was recorded on their clinical record, and participants were followed up until the earliest of: an outcome of interest; death; end of follow-up (29 February 2016); or last registration date with their GP practice. Participants were removed from our sample if they were less than 40 years of age, had less than 12 months of "research quality" data, were simultaneously prescribed more than one lipid-regulating agent (due to the difficult of assigning these to a single exposure group), or were diagnosed with dementia before or on the date of the index event.

@@

We considered seven lipid-regulating drug classes based on groupings in the British National Formulary (BNF)[@wishart2017], namely: statins, fibrates, bile acid sequestrants, ezetimibe, nicotinic acid groups, ezetimibe and statin (representing one treatment containing both drugs, rather than the two classes being prescribed concurrently), and omega-3 fatty acid groups.

A participant’s drug class was assigned based on their first recorded prescription, and any drug switching was ignored in an effort to mimic an intention-to-treat approach. We did however examine how often the initial drug class was stopped (defined as last prescription of the primary class being followed by at least six months of observation), added to (defined as a second drug class being prescribed before the last prescription of the initial class), or switched (defined as a second drug class being prescribed after the last prescription of the initial class).

@@

### Outcomes

<!--- Include short bit on follow-up here, describing how and why it was defined --->

We considered five outcomes as part of this analysis: probable  Alzheimer’s disease, possible Alzheimer’s disease, vascular dementia, other dementia, and a composite all-cause dementia outcome (Supplementary Figure 1). When two or more outcomes were coded in a participant’s clinical record, a decision tree was used to differentiate between them (Supplementary Figure 1). The diagnosis date of the outcome was determined by the first record of a relevant code.

@@

A range of additional variables were included in the analysis, intended to address differential distributions of potnetial confounding variables between those who had a

<!--- Need a lot more here - see other thesis for example of how covariates are described --->

The analysis was adjusted for a range of baseline covariates including sex, grouped year of entry into the cohort (<2000, 2000-2004, 2005-2009, >2010),

Charlson co-morbidity index,

Index of Multiple Deprivation (IMD),

consultation rate,

alcohol (current, former,  never),

smoking (current, former,  never),

BMI,

baseline total cholesterol,

and history of cardiovascular disease, coronary bypass surgery, coronary artery disease, peripheral arterial disease, hypertension, chronic kidney disease, and Type 1 and Type 2 diabetes.

All covariates were determined at index and definitions for each can be found in Supplementary Table 1.

@@

Missing data was handled using a multiple imputation approach. Variables with missing observations were identified, and

Binary variables were modelled using a

continuo

As per best practice, all other variables, including the out

The MICE command in Stata was used

In Cox models, there is the option to

Age vs time-since entry

Expand on use of age as the time scale, but describe the problems this introduced

@@

Potential biases included time varying confounding, selection bias due to censoring on death and 
We use a Cox proportional hazards model with a time-varying treatment indicator.

Expand on post estimation analyses, which showed that nothing except age had an effect.

$$h(t) = h_o(t) \times exp(b_1x_1 + b_2x_2 + ... +b_px_p)$$

@@

__Bit on why Cox models are slightly different from others, due to proportional hazards assumptions__

Hazard rations are

Test for non-proportionality are therefore very important, as models which fail

Problems with testing for non-proportional hazards when

<!----------------------------------------------------------------------------->
(ref:cprd-dag-1-cap) __Directed acyclic graph showing assumed relationship between exposure and covariate:__

(ref:cprd-dag-1-scap) Directed acyclic graph showing assumed relationship between exposure and covariate

```{r cprd-dag-1, echo = FALSE, results="asis", fig.cap='(ref:cprd-dag-1-cap)', out.width='100%', fig.scap='(ref:cprd-dag-1-scap)'}
knitr::include_graphics(file.path("figures/sys-rev/prisma_flow.png"))
```
<!----------------------------------------------------------------------------->

### Immortal time bias and time-varying treatment indicators

Immortal time bias is a type of misclassification bias. It occurs when

One potential way to address this is the use of a time-varying treatment indicator, a factor which modi

In the case of this analysis, this treatment indicator is statin use. Using codes for statins (see Appendix \@ref())

<!----------------------------------------------------------------------------->

### Estimating the value of the time-varying confounders

Mean time from index event to first prescription of statins was 2.4 years. This negates the promised benefit of ruling out confounding by indication (where the test result leads to the prescription of the treatment and also increases the risk of the outcome, distorting the relationship between the two), as there is no relationship between index TC/LDL-c and eventual LRA prescription.

Additionally, the time between index event and prescription does lead to a problem in terms of time varying confounding, as an average time of 2.4 years between current measurement of the covariates and treatment switching means there is plenty of time for the value of the covariate to change. This is problematic when the decision to change treatments (in this case to move from no LRA use to LRA use) is influenced by a set of prognostic factors that in turn may have been influenced by the initial treatment decision, as is likely to be the case for a range of covariates included in the model. For example:

    _No CVD (t=0) -> No LRA (t=0) -> CVD (t=1) -> LRA (t=1) -> Dementia (t=2)_

In this case, the decision to move to LRA use is influenced by CVD status at _Time 1_, which will not be captured by adjusting only for CVD status at _Time 0_.

In practice, this means that the value of the prognostic factor should be regularly captured

However, in electronic health records, a change in the value of the prongostic factors is only important if it is recorded in a patients record, as for it to have an impact on treatment decisions, it must be recorded.

This means we can find the most recent value of the covariate before the switch and apply a marginal structural model approach, filling all values for that variable before the most recent measure with the baseline measurement, and all after the most recent meausure with the value of the most recent measure (on the basis that you won't go from having CVD back to not having CVD).

i.e.  
Timepoint 12345678  
CVD       00001111  
Treatment 00000111

Split into 3 month blocks since index event and use the same approach as above to work out the values of each covariate at each time point.

Note: this will be harder for things that are not dichotomous and can go up as well as down. Examples include total cholesterol and BMI, which can go up as well as down.

<!----------------------------------------------------------------------------->

### The effect of total cholesterol or LDL-cholesterol on LRA prescription

It would be fair to assume that the baseline total cholesterol/LDL-cholesterol would at least in part predict the likelihood of someone being prescribed a statin.

However, this is not the case. Baseline cholesterol level are predicted to be a poorer instrument for than QRISK2 score,[@hippisley-cox2008] which estimates a patients' 10-year risk of a cardiovascular event. Current NICE guidelines state that those with a QRISK score of 10% or higher, and in whom lifestyle modification is ineffective/inappropriate, should received a lipid regulation agent. However, this analysis could not find any effect of QRISK2 scores on statins prescription levels at 6 months.

>>>>

```{r, echo = FALSE, warning=FALSE, message=FALSE}
source("R/helper.R")
knitr::read_chunk("R/04-Code-CPRD-Analysis.R")
doc_type <- knitr::opts_knit$get('rmarkdown.pandoc.to') # Info on knitting format
```

```{r load-files, echo = FALSE}
```

<!-- ## Additional ideas -->

<!-- TODO Need to add some information on the generation and use of propensity scores -->

<!-- One of the problems I'll point out with the trials in the systematic review chapter (Chapter \@ref(sys-rev-heading)) is that they only included people with a high cardiovascular risk - we are kind of doing the same by using elevated test cholesterol results as the index event. -->

<!-- Might be able to get around this by pointing out that you need some index event to be able to do this analysis (?) -->

<!-- __Things we tried__ -->

<!-- - Conditioning entry to the cohort on QRISK score (as defined by codes) rather than by . Need to be able to demonstrate here that not many meeting our original entry criteria actually go on to have a statin quickly. Need total number of those starting and average time to start. -->
<!-- - Positive controls. Need to be able to explain why these controls have such wildly increased results. Cross-reference these results with the Hernan paper. -->
<!-- - Competing risk analysis, with death as a competing risk. Need numbers of deaths in each group, and preliminary results -->
<!-- - Allowing for time-varying confounders for binary covariates. -->

<!-- __Things we could try__ -->

<!-- - Work through data using positive control to see if the way it is being added is what is causing the issues. -->
<!-- - Examine Cramer et al. thesis and see what they did -->
<!-- - Marginal structural models approach (a la J. Sterne) -->

<!--------------------------------------------------------------------------->

<!-- TODO Check lipid-regulating vs lipid regulating -->

<!-- TODO Check all fig/table cross-references evaluate -->

## Lay summary

Electronic health record (EHR) databases are large collections of patients data, used to manage patient administration and care. Under these systems, whenever a patient attends their GP, their clinical data is recorded in a central database using a standard set of codes. These databases have several advantages over traditional methods of data collection, including the number of people they contain and the length of time for which participants are followed. This is particularly important when studying diseases such as dementia, which may begin to develop in patients long before symptoms are seen.

This analysis makes use of the Clinical Practice Research Datalink (CPRD), which contains the electronic medical records of more than 3 million people from general practices across the UK. Using this data, the analysis presented in this chapter examined the effect of treatment which lower cholesterol levels, such as statins, on the risk of dementia and related outcomes.

Little evidence for an effect of lipid regulating agents effect on the risk of Alzheimer's disease was found, with the exception of a slightly increased risk in those prescribed a certain type of drugs called fibrates. In contrast, I found an increased risk of vascular and other (i.e. non-Alzheimer's) dementia with lipid regulating agent use.

This increased risk is unexpected, and is very likely to be due to the presence of bias in the analysis. This bias, called "confounding by indication", is caused when those who are prescribed a statin are more at risk of vascular dementia for a range of reasons, and so it appears as if statins are harmful. However, despite this limitation, the analysis presented provides an important source of information which will be used in later chapters.

<!--------------------------------------------------------------------------->

@@

In this Chapter, I present the results of an analysis of a large population-based electronic health record dataset to investigate the relationship between lipid regulating agent (LRA) use and dementia outcomes.

The analysis aims to address important two limitations of the current evidence base as identified by the systematic review presented in Chapter \@ref(sis-rev-heading).

Firstly, it explicitly examines vascular dementia as an outcome. The systematic review presented in the previous chapter identified an evidence gap around the effect of lipid regulating agents on the risk of vascular dementia. As triangulation exercises require as many diverse sources of evidence as possible, this analysis provides a source of information on this outcome.

Secondly, and in a similar vein, the analysis intentionally takes a different analytical approach to that most commonly used to examine the effect of statins on dementia (as identified by the systematic review). Specifically, this involved a concerted effort to address immortal time bias though use of a Cox Proportional time-varying treatment indicator. This approach provides a evidence source at risk of a distinct set of biases due to the alternative analytical strategy that will be incorporated into the triangulation exercise presented in Chapter \@ref(discussion-heading).

<!--------------------------------------------------------------------------->

@@

### Study protocol

An *a priori* protocol for this study was published, [@walker2016] and amendments to this are recorded in Appendix \@ref(appendix-cprd-amendments).[@vonelm2008]

<!-- QUESTION If protocol was written by Venexia, how do I reference this? -->

<!-- TODO Add amendments to Appendix -->

<!--------------------------------------------------------------------------->

@@

Previously known as the General Practice Research Database (GPRD), the Clinical Practice Research Datalink (CPRD) is a large population-based, electronic health record (EHR) database.[@herrett2015] The database has been collecting primary care data from participating practices across England since 1987.[@williams2012; @wood2001revitalizing] It contains the primary care records for more than 10 million primary care patients in England, and is broadly representative of the UK population in terms of age, sex and ethnicity.[@herrett2015; @mathur2014]

To avoid the ambiguity of interpreting free-text clinical notes and to allow for easy analysis of the resulting data, the CPRD primarily collects data using a predefined coding system known as Read codes.[@booth1994] All clinical events, included clinical test results and diagnoses, can be identified by a specific Read code. The codes use a nested approach (see Table \@ref(tab:readExample-table)), with the initial characters defining broad diagnostic topics (e.g. Eu... - Mental and behavioural disorders), while subsequent characters provide additional information on the specific condition diagnosed (e.g. Eu001 - Dementia in Alzheimer's disease with late onset).

@@

<!--------------------------------------------------------------------------->
(ref:readExample-caption) Example of CPRD Read code hierarchy, showing how "Dementia in Alzheimer’s disease with late onset" (_Eu001_) is nested under the top-level of "Mental disorders" (_Eu..._). Broad topics are specified using the initial two alpha-numeric characters of the Read code, while subsequent characters are used to define specific conditions and context.

(ref:readExample-scaption) Example of CPRD Read code hierarchy

```{r readExample-table, message=FALSE, results="asis", echo = FALSE}
```
<!--------------------------------------------------------------------------->

Lists of relevant codes for each of the index events, exposures and outcomes used in the analysis were created. Each of these are described in more detail in the following sections.

### Cohort definition

This analysis included all participants registered at a participating practice between 1 January 1995 and 29 February 2016 who had a flag for "research quality" data. Records pre-dating the 1995 cutoff were included in the original CPRD extract obtained for this analysis. However, these were excluded from the analysis as data quality and reliability is thought to be higher after this date.[@wolf2019] Additionally, individuals with less than 12 months of continuous records prior to cohort entry were excluded, making the effective start date of the cohort 1 January 1996.

Participants were included in the study cohort if their record contained any of the following index events: a Read code for a diagnosis of hypercholesterolemia or related condition; a Read code for prescription of a lipid-regulating agent (such as statins); a total cholesterol test result of >4 mmol/L; or an LDL-c test result of >2 mmol/L. These index events allowed me to define a population of participants who were either at risk of hypercholesterolemia, as indicated by the elevated total or LDL cholesterol test results, or had already been diagnosed with it, as indicated by a diagnostic code/related prescription.

All events of interest were identified using predetermined code lists, which are available for inspection from the archived repository accompanying this analysis (data/code availability is discussed in Section \@ref(cprd-data-avail)).

The index date for a participant was defined as the date where the first relevant code or test result was recorded on their clinical record, and participants were followed up until the earliest of: an outcome of interest; death; end of follow-up (29 February 2016); or last registration date with their GP practice. Participants were removed from the sample if they were less than 40 years of age, had less than 12 months of "research quality" data, were simultaneously prescribed more than one lipid-regulating agent (due to the difficult of assigning these to a single exposure group), or were diagnosed with an outcome of interest before or on the date of the index event (i.e. had less than one full day of follow-up).

<!--------------------------------------------------------------------------->

@@

We considered seven lipid-regulating drug classes based on groupings in the British National Formulary (BNF),[@wishart2017] namely: statins, fibrates, bile acid sequestrants, ezetimibe, nicotinic acid groups, ezetimibe and statin (representing one treatment containing both drugs, rather than the two classes being prescribed concurrently), and omega-3 fatty acid groups.

A participant’s drug class was assigned based on their first recorded prescription, and any drug switching was ignored in an effort to mimic an intention-to-treat approach. We did however examine how often the initial drug class altered according to one of three criteria:

- __stopped__: defined as the last prescription of the primary class being followed by at least six months of observation; 
- __added__: defined as a second drug class being prescribed before the last prescription of the initial class; and 
- __switched__: defined as a second drug class being prescribed after the last prescription of the initial class.

<!--------------------------------------------------------------------------->

@@

### Outcomes {#cprd-outcomes}

We considered five outcomes as part of this analysis: probable  Alzheimer’s disease, possible Alzheimer’s disease, vascular dementia, other dementia, and a composite all-cause dementia outcome. When two or more outcomes were coded in a participant’s clinical record, a decision tree was used to differentiate between them (see Figure \@ref(fig:decisionTreeFig)). The diagnosis date of the outcome was determined by the first record of a relevant code.

<!----------------------------------------------------------------------------->
(ref:decisionTreeFig-cap) Decision tree for assigning dementia subtypes, based on the presence of Read codes in the patient's record. Note that an outcome of "Probable" or "Possible" Alzheimer's disease (AD) requires the absence of any vascular outcome codes.

(ref:decisionTreeFig-scap) Decision tree for assigning dementia subtypes

```{r decisionTree, include = FALSE}
```

```{r decisionTreeFig, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:decisionTreeFig-cap)', out.width='100%', fig.scap='(ref:decisionTreeFig-scap)'}
knitr::include_graphics(file.path("figures","cprd-analysis","decision_tree.png"))
```
<!----------------------------------------------------------------------------->

Participants were censored at the earliest of date of: diagnosis; death; transfer out of the study cohort; or end of follow-up (29th February 2016).

@@

A range of additional variables were included in the analysis, intended to address the different distributions of potential confounding variables between those who were prescribed an lipid-regulating agent and those who were not.

Demographic covariates adjusted for included age and gender. Age was calculated at date of entry into the cohort. Socioeconomic status was proxied using the Index of Multiple Deprivation (IMD) 2010, which draws on seven domains (income; employment; education, skills and training; health and disability; crime; barriers to housing and services; living environment) to create an overall deprivation score for each of 32844 statistical geography areas in England. To help preserve patient privacy, IMD score is only available from the CPRD in twentiles, with 1 indicating the least deprived and 20 indicating the most deprived.  Smoking and alcohol use was determined at index, and participants were categorised as current, former, or never users of each.

Body mass index (a summary measure calculated as $weight/height^2$), baseline total cholesterol and baseline LDL cholesterol measures were obtained, using the last recorded value prior to the index date. A variable indicating grouped year of entry into the cohort (<2000, 2000-2004, 2005-2009, >2010) was included to allow for changes in prescribing trends across the lifetime of the cohort. To assess healthcare utilisation, I adjusted for the average annual number of consultations between the beginning of a patients data and their entry into the cohort.

Finally, presence of a range of related conditions at baseline were accounted for, including cardiovascular disease, coronary bypass surgery, coronary artery disease, peripheral arterial disease, hypertension, chronic kidney disease, and Type 1 and Type 2 diabetes. In addition to adjusting for these covariates individually, a Charlson co-morbidity index (CCI) score was calculated for each participant. The CCI is a weighted index that uses presence and severity of a number of conditions to enable adjustment for the general health of a participant in terms of their mortality risk.[@charlson1987new] Inclusion of this index allowed me to attempt to adjust for the general health of patients included in the analysis.

Codelists for all covariates can be found in the archived data repository accompanying this analysis (see Section \@ref(cprd-data-avail)).

<!--------------------------------------------------------------------------->

@@

Missing data are a recognised issue in electronic health records databases,[@wells2013strategies] given that they contain administrative data, collected primarily for the purposes of patient management and care rather than academic research.

<!-- IDEA Add more detail here on the missing data approach -->

In this analysis, missing data were handled using a multiple imputation approach. Variables with missing observations were identified, and 20 imputed datasets were created.[@sterne2009] Nominal variables with missing values were modelled using multinomial logistic regression, while continuous variables were modelled using linear regression. As per best practice, all variables used in the analytic model, including the outcome, were included in the imputation model.[@moons2006using] Imputation was performed using the MICE (Multiple Imputation by Chained Equations) command in STATA16.

Missing data was only considered an issue for variables where a numerical test result was expected (e.g. BMI), or where a code existed for the absence of the condition (e.g. categorical smoking status). This approach was necessary, as absence of a code for other treatments or conditions (e.g. statin use or dementia) was assumed to indicate absence of the treatment/condition, rather than being considered missing.[@wells2013strategies]

Using multiple imputation to handle missing data is an alternative to a "complete case" approach,[@pigott2001review] where participants missing any covariate are dropped from the dataset. As a sensitivity analysis, I preformed and compared the results of both methods,[@hughes2019] to investigate the impact of multiple imputation on the results.

&nbsp;
<!--------------------------------------------------------------------------->

@@

A Cox proportional hazards (PR) model was used to estimate the effect of statins on dementia outcomes. Cox PR models are defined, in general terms, as:

\begin{equation}
  h(t) = h_o(t) \times exp(b_1x_1 + b_2x_2 + ... +b_px_p)
  (\#eq:cox-model)
\end{equation}

@@

A Cox PR model was chosen for this analysis as it inherently accounts for the length of time participants spend in each exposure group. Using this approach, time-at-risk can be properly attributed, thus mitigating the impact of immortal time bias. This is discussed in detail in the following section.

<!-- IDEA Bit on why Cox models are slightly different from others, due to proportional hazards assumptions. Test for non-proportionality are therefore very important, as models which fail-->

<!-- IDEA Need to describe accounting for clustering here: id it somethings I did? See other thesis for clustering by practice -->

<!----------------------------------------------------------------------------->
&nbsp;

### Immortal time bias and time-varying treatment indicators {#cprd-immortal-time-bias}

Immortal time bias covers two distinct but related types of bias. The first presentation, the selection bias aspect (Panel A, Figure \@ref(fig:immortalTimeBias)), occurs when time prior to the exposure is excluded leading to the exposed and control groups being followed up from different time points.[@levesque2010] For example, if the unexposed group are followed from a cholesterol test result, while the exposed group is followed from date of LRA prescription, any events that occur in the exposed group prior to the exposure event will be inappropriately excluded from the analysis.

The second presentation of immortal time bias is as a type of misclassification bias (Panel B, Figure \@ref(fig:immortalTimeBias)). It occurs when the exposure time prior to the exposure date, and any events occurring within it, is inappropriately assigned to the exposed group. This second presentation appears to be common in the existing literature, as several of the studies included in the systematic review presented in Chapter \@ref(sys-rev-heading) were identified as being at risk of immortal time bias following formal risk of bias assessment using the ROBINS-I tool (see Section \@ref(risk-of-bias-subheading)).

This analysis is primarily concerned with the second presentation of immortal time bias, as all participants were followed from a common index date (earliest of: date of raised cholesterol test results; hypercholesterolemeia diagnosis; or LRA prescription). To address the potential for this second form of immortal time bias in the analysis, I employed a time-varying indicator of treatment status to correctly allocate time-at-risk to the exposed and unexposed groups.[@levesque2010]

Under this approach, all patients start in the unexposed group, & contribute time-at-risk until they are prescribed a lipid regulating agent and move into the exposed group. Note, patients for whom prescription of a lipid regulating agent was the index event only contribute time to the exposed group (i.e. they enter the cohort and move into the exposed group on the same day).

<!--------------------------------------------------------------------------->
(ref:immortalTimeBias-cap) Diagram illustrating the two presentations of immortal time bias, as a selection bias (Panel A) and a misclassification bias (Panal B).

(ref:immortalTimeBias-scap) shortcap

```{r, immortalTime, include = FALSE}
```

```{r immortalTimeBias, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:immortalTimeBias-cap)', out.width='100%', fig.scap='(ref:immortalTimeBias-scap)'}
knitr::include_graphics(file.path("figures","cprd-analysis","immortal_time.png"))
```
<!--------------------------------------------------------------------------->

<!-- IDEA Finish paragraph from fragments below -->

<!-- In Cox models, there is the option to -->

<!-- Age vs time-since entry -->

<!-- Expand on use of age as the time scale, but describe the problems this introduced -->

As part of a Cox proportional hazard model, there is the option to use either absolute time in cohort or participants age as the time scale of interest.[@lamarca1998; @gail2009; @pencina2007] A model using age as the time axis inherently accounts, or adjusts, for participants age as a potential confounder of the exposure-outcome relationship. As such, the main analyses presented all used age as the time axis.

<!-- TODO Will need to expand on choice of time axis here -->

<!-- IDEA In order to explore whether the choice of time axis affects the results of the analysis,I ran two simple models, one using participants age as the time-scale and a second using absolute time-since-entry and adjusting for participant age. -->

&nbsp;
<!--------------------------------------------------------------------------->

### Sensitivity analyses

The primary analysis examined the effect of a lipid regulating agent on dementia risk, stratified by outcome and drug class. To assess the robustness of the results, a number of sensitivity analyses were performed. These are described in the following sections.

#### Control outcomes

In addition to the primary outcomes of interest (described in Section \@ref(cprd-outcomes)), I extracted data on two additional control outcomes. The inclusion of control outcomes in observational analyses are a useful technique to assess the strength of uncontrolled confounding,[@lipsitch2010] and these outcomes are usually class as either "negative" or "positive" outcomes. Negative outcomes are those without a likely causal path between the exposure and outcome (see Figure \@ref(fig:negativeOutcome) for a directed acyclic graph describing an ideal negative outcome).

Conversely, positive control outcomes are those with a known causal association with the exposure of interest, ideally sourced from large well conducted randomised controlled trials. Positive control outcomes are useful in observational epidemiology, as if the analysis can reproduce a known result for the control outcome, confidence in the result for the outcome of interest is increased.

Due to the wealth of data available on statins as a lipid regulating agent, control outcomes were chosen in reference to this drug class. Muscular backpain was chosen as a negative control outcome in this analysis. Despite observational analyses suggesting a link between statins and muscle pain (as opposed to more serious complications such as myopathy),[@selva-ocallaghan2018] systematic reviews of the adverse events of statin use[@collins2016a] and N-of-1 trials explicitly exploring the association of statin use with muscle pain[@herrett2021] have found little evidence supporting an effect. As such, if statins are not associated with backpain in this analysis, then confidence in the results for the dementia outcomes is increased.

Additionally, incident ischemic heart disease was included as a positive control outcome, given the well-established protective effect of lipid-lowering treatment, via statins, on the risk of this condition.[@collins2016a] Similar to the backpain outcome, if the analysis strategy can reproduce this known protective association, this will provide evidence that potential confounders have been sufficiently adjusted for.

<!-- QUESTION - should those with baseline cardiovascular conditions be excluded from the IHD/MI analysis? -->

<!-- TODO Need to add bit to this section and to results on outcome for Type 2 Diabetes outcome -->

<!----------------------------------------------------------------------------->
(ref:negativeOutcome-cap) Causal diagram showing relationship between exposure $X$, outcome $Y$, confounders (measured $C$ and unmeasured $U$) and an ideal negative outcome $N$. Note the absence of any arrow between $X$ and $N$. In this scenario, any association observed between $X$ and $N$ is due to the presence of uncontrolled confounders $U$ (assuming $C$ has been adjusted for).

(ref:negativeOutcome-scap) DAG of ideal negative outcome

```{r negativeOutcome, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:negativeOutcome-cap)', out.width='100%', fig.scap='(ref:negativeOutcome-scap)', message=FALSE, warning = FALSE}
library(dagitty)
library(ggdag)

ggplot2::ggsave(filename = file.path("figures","cprd-analysis","negativeOutcome.png"),
  dagitty::dagitty("dag {
    U -> N
    C -> N
    X <- U -> Y
    X <- C -> Y
    X -> Y
    bb =\"-2.368,-2.972,2.597,2.726\"
    C [pos=\"-0.3,0\"]
    N [pos=\"0.5,1\"]
    U [pos=\"-0.15,1.5\"]
    X [exposure,pos=\"0,-1\"]
    Y [outcome,pos=\"0.5,-1\"]
  }"
) %>% ggdag_classic() +
  theme_dag()

,
height = 3
)

knitr::include_graphics(file.path("figures","cprd-analysis","negativeOutcome.png"))

```
<!----------------------------------------------------------------------------->

#### Impact of additional covariates

To observe the effect of adjusting for additional covariates, I ran a model unadjusted except for age (captured via the time axis in the Cox model) and gender, and compared the results with the full adjusted model.

#### Sensitivity cohorts

Two sensitivity cohorts were also created. The first stratified by year of entry into the cohort in an attempt to assess for time period effects. The second removed participants who may have been pregnant  (coded as under 55 and female) to assess the robustness of the estimates, as statins are contraindicated in pregnancy, .[@karalis2016]

<!-- IDEA Make sure to include sex as a sensitivity analysis, given that that is the purpose of the IPD analysis -->

#### Statin properties

<!-- IDEA Could frame this as providing evidence that the effect is not due to statins acting on lipid levels, as then would not expect difference (unless all very hard acting statins are in the lipophilic type) -->

As detailed in the introduction, the properties of statins may be important in their effect, based on the ability of lipophilic statins to cross the blood brain barrier (see Section \@ref(intro-statins) in the Introductory Chapter).

To explore whether any observed associations in the statin analysis varied by statin property, a sensitivity analysis was performed, stratifying by statin lipophilicity.

<!-- TODO Previous review of the evidence have recommended that future studies take account of statins varying ability to cross the BBB [@shepardson2011] -->

#### Impact of dementia code lists

As part of an exploratory analysis of the effect of the choice of code lists on the analysis, I created an alternative Alzheimer's disease and non-Alzheimer's dementia outcome using code lists from a previous study published in 2009 by Smeeth _et al_.[@smeeth2009]

This previous study used a propensity matching approach to estimate the association of statins with a range of outcomes, and found evidence for a protective effect of statin use on all-cause dementia (`r estimate(0.81, 0.69,0.96,type = "HR")`) and non-AD dementia (`r estimate(0.82,0.69, 0.97,type = "HR")`), but little evidence of an effect on AD (`r estimate(0.81,0.49, 1.35, type="HR")`).

The code lists used were obtained through correspondence with the authors of that study, and are available for inspection (see Section \@ref(cprd-data-avail)).

## Results

### Patient characteristics

```{r characteristics, echo = FALSE}
```

```{r attritionFigure, include = FALSE}
```

Of the `r comma(attrition[1,1])` participants included in the extract, `r comma(attrition[14,1])` met the inclusion criteria (Figure \@ref(fig:cprdFlowchart)), with a total follow-up of `r comma(total_followup)` patient years at risk.

<!--------------------------------------------------------------------------->
(ref:cprdFlowchart-cap) Attrition of CPRD participants as the eligibility criteria were applied. Most attrition was due to the absence of an index event of interest.

(ref:cprdFlowchart-scap) Attrition of CPRD participants

<!-- This figure needs to have comma seperated numbers in order to agree with text. -->

```{r cprdFlowchart, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:cprdFlowchart-cap)', out.width='100%', fig.scap='(ref:cprdFlowchart-scap)'}
knitr::include_graphics(file.path("figures","cprd-analysis","cohort_attrition.png"))
```
<!--------------------------------------------------------------------------->

The median participant age at index was `r age_text` and participants were followed up for a median of `r fu_text`. During follow-up, an all-cause dementia diagnosis was recorded for `r comma(p1[1,2])` patients  (`r comma(p1[2,2])` probable AD, `r comma(p1[3,2])` possible AD, `r comma(p1[4,2])` vascular dementia, `r comma(p1[5,2])` other dementia: Table \@ref(tab:followUp-table)). The distribution of baseline characteristics across the seven drug classes can be seen in Table \@ref(tab:cprdCharacteristics-table).

\blandscape
<!--------------------------------------------------------------------------->
(ref:cprdCharacteristics-caption) Patient characteristics by drug class. Summary statistics are presented as "% (N)" unless otherwise specified in the variable name.

(ref:cprdCharacteristics-scaption) Patient characteristics by drug class

```{r cprdCharacteristics-table, message=FALSE, results="asis", echo = FALSE}
```
<!--------------------------------------------------------------------------->
\elandscape

A substantial majority (`r percentage.statins`) of participants prescribed a lipid-regulating agent were prescribed a statin. I excluded the "Ezetimibe and statins" and "Nicotinic acid groups" classes from subsequent analysis based on the extremely small number of participants in these groups (N=127 & N=168 respectively; see Table \@ref(tab:cprdCharacteristics-table)). Note that the "Ezetimibe and statins" treatment group represent those prescribed a single treatment containing both ezetimibe and statins, rather than those where the two treatments were prescribed concurrently.

<!-- QUESTION Describe impact of SSA table - is the fact that many people stopped statin treatment (~10%) an issue? Comment specifically in the context of a -->

<!-- IDEA Described the average length of time between patients being prescribed a statin and being diagnosed with dementia -->

<!--------------------------------------------------------------------------->
(ref:cprdSSA-caption) Participants who stopped, switched or added treatments by initial treatment type.

(ref:cprdSSA-scaption) Participants who stopped, switched or added treatements by initial treatment type

```{r cprdSSA-table, message=FALSE, results="asis", echo = FALSE}
```
<!--------------------------------------------------------------------------->

The stopping, addition and switching of drug classes was common across all drug classes (Table \@ref(tab:cprdSSA-table)).

<!--------------------------------------------------------------------------->

```{r missing-data, echo = FALSE}
```

```{r azd-text, echo = FALSE}
```

Full covariate information was available for `r missingtext`. Six key variables had some missing data: IMD 2010 score was missing for `r imdtext`, because it is only recorded for English practices; alcohol status was missing for `r alcoholtext`; smoking status was missing for `r smokingtext`; BMI, or a calculated BMI from height and weight measurements, was missing for `r bmitext`; baseline total cholesterol was missing for `r choltext`; and baseline LDL cholesterol was missing for `r ldltext`.

### Primary analysis
